Schema Evolution in Interactive Programming Systems

### 2024-08-24 paper.tex
Reviewer A nitpicks

Many improvements to the practice of programming are due to speeding up feedback loops. Replacing punched cards with interactive terminals was one of the greatest leaps of productivity in the history of programming. Unit tests and interactive type checking provide faster feedback on errors that modern programmers now expect. Research on Live Programming~\cite{rein2018exploratory} seeks to provide immediate feedback on runtime behavior while editing. Feedback is equally valuable in collaborative programming~\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) to convey changes between coworkers with less friction.
However there is still much room to improve feedback loops throughout the software development enterprise.

In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development. While Schema Change has been extensively studied in the context of databases~\cite{erhard06}, we observe that analogous problems occur in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in \S\ref{sec:related} Related Work. In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also types and specifications of all sorts, whether expressed in a formal language or left implicit in code.

Our primary contribution is a suite of challenge problems exemplifying a diverse range of typical schema change scenarios that interrupt feedback loops. We believe these problems have not been completely solved in theory or practice (see \S\ref{sec:related} Related Work). While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches. None of our ongoing efforts yet address the full range of problems. Indeed the motivation of this paper is to provide a basis for communicating and collaborating between related efforts, not to mention motivating further research efforts. We hope these challange problems serve as examples to ground future research and as a basis for evaluating the results. We invite contributions of new challenge problems and look forward to discussing competing solutions.

@@

\paragraph{Challenge Problems.}
These dimensions, layers, and variants offer a wide range of configurations for challenge problems, but
our aim is not to enumerate all of them. We use them to recast a number of
existing challenges in real-world programming systems as instances of the same problem of
schema evolution. The case studies show how well-known solutions or problems in one domain
map to another domain and they also provide and inspiration and a benchmark for
developers of future programming systems.

@@

\subsection*{Remarks: Requirements and Implementation}
The key requirement from the live programming perspective is to ``do minimal harm''.
The migration of \emph{data} needs to be done automatically and the system should strive
to produce a new state that is as near as possible to the previous state. Migrating
\emph{code} can be done in various ways,
but it is desirable to avoid breaking the \texttt{render} function as this would make the
new application state impossible to visualize.

@@

The \emph{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers,
and 3) reference these identifiers from the orders. Our example took the simple approach of
merging entities whose attributes are all equal. But that might not be correct in all cases.
What if there was a typo in one of the addresses that made it look different from other instances
of the customer? To repair that error we need to merge the falsely distinct customers into one
and fix any references from orders. We call this operation \emph{Merge Entities} -- note that
it is not a schema change but a data change, yet nevertheless not commonly supported in data APIs.

@@

\noindent
In database systems, the system can use a mechanism such as a
view and execute the original query over the view. When applying code migrations,
the system should minimize the need for users to manaually intervene. It should also
minimize the possibilities that developer-written code is subtly incorrect in edge cases,
preferably by reducing the need for developer-written code.

@@

This challenge invites a change of perspective in both live programming and local-first software.
Live programming must move from being a solitary activity to a collaborative workflow, and one
where the collaboration is not just on editing source code but on live integrated code and data,
as in the Smalltalk/Lisp images of old.
%\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!}
Local-first software must move from automatically converging data replicas
to also interactively managing long-term divergence. That implies either application state
is being directly exposed to the user, or the application code is using an API to present
version control affordances. We encourage both communities to think outside the box of Git,
which has proven to baffle not only end-users but also a substantial fraction of developers.

@@

Although there is likely no silver bullet to navigate these tradeoffs, a good solution to this
problem would give developers or users tools to manage these tradeoffs. For example, a system
might allow developers to specify the desired behavior for a particular pair of schemas based
on the requirements of the domain. Or a system might even allow users to manually
disambiguate their intent for a given write.

>>>>

Many improvements to the practice of programming have come from speeding up feedback loops. Replacing punched cards with interactive terminals was one of the greatest leaps of productivity in the history of programming. Unit tests and interactive type checking provide faster feedback on errors that modern programmers now expect. Research on Live Programming~\cite{rein2018exploratory} seeks to provide immediate feedback on runtime behavior while editing. Feedback is equally valuable in collaborative programming~\cite{ProGit, goldman2011real,kurniawan2015coder,replit}) to convey changes between coworkers with less friction.
However there is still much room to improve feedback loops throughout software development.

In this paper we identify a common problem, \emph{Schema Change}, that stands athwart feedback loops in diverse contexts of software development. Schema Change has been extensively studied in the context of databases where it necesitates manual and error-prone data migrations.~\cite{erhard06} We observe analogous problems occuring in varied and less-studied contexts. For example, many live programming techniques treat state as ephemeral and recreate it after every edit, but when the shape of longer-lived state changes then the illusion of liveness is shattered -- hot reloading works until it doesn't. Likewise collaborative programming works smoothly when code changes can be exchanged that do not alter the expected shape of external data, but changes that cross that line must be coordinated outside of the version control system. We discuss further examples in \S\ref{sec:related} Related Work. In general, when the shape of data becomes a dependency then changes to that data's shape can require manual intervention which interrupts feedback loops. To coordinate research on these related problems we shift the focus to that of \emph{Interactive Programming Systems}~\cite{techdims} of all sorts and generalize the term `schema' to include not just database schema but also data types and specifications of all sorts, whether expressed in a formal language or left implicit in code.

Our primary contribution is a suite of challenge problems exemplifying a diverse range of typical schema change scenarios that interrupt feedback loops. We believe these problems have not been completely solved in theory or practice (see \S\ref{sec:related} Related Work). While some of these problems come from our individual research efforts we have attempted to make neutral problem statements not slanted to our preferred approaches. None of our ongoing efforts yet address the full range of problems. Indeed the motivation of this paper is to provide a basis for communicating and collaborating between such related efforts, as well as to facilitate further research efforts. We hope these challange problems serve as examples to ground future research and as a basis for evaluating its results. We invite contributions of new challenge problems and look forward to discussing competing solutions.

@@

\paragraph{Challenge Problems.}
These dimensions, layers, and variants generate a wide range of configurations for challenge problems, but
our aim is not to cover all of them. Instead we use them to recast a number of
existing challenges in real-world programming systems as instances of the same problem of
schema evolution. The case studies show how well-known solutions or problems in one domain
map to another domain and they also provide and inspiration and a benchmark for
developers of future programming systems.

@@

\subsection*{Remarks: Requirements and Implementation}
The key requirement from the live programming perspective is to ``do minimal harm'' to the experience of liveness.
The migration of \emph{data} needs to be done automatically and the system should strive
to produce a new state that is as near as possible to the previous state. Migrating
\emph{code} can be done in various ways,
but it is desirable to avoid breaking the \texttt{render} function as this would make the
new application state impossible to visualize.

@@

The \emph{Extract Entity} operation must 1) merge duplicates, 2) assign unique identifiers,
and 3) reference these identifiers from the orders. Our example took the simple approach of
merging entities whose attributes are all equal. But that might not be correct in all cases.
What if there was a typo in one of the addresses that made it look different from other instances
of the customer? To repair that error we need to merge the falsely distinct customers into one
and fix any references from orders. We call this operation \emph{Merge Entities} -- note that
it is not a schema change but a data change, yet nevertheless not commonly supported as an operaton in database APIs or DMLs.

@@

\noindent
In database systems, the system can use a mechanism such as a
view and execute the original query over the view. When applying code migrations,
the system should minimize the need for users to manually intervene, which interrupts feedback from the migration. It should also
minimize the possibilities that developer-written code is subtly incorrect in edge cases,
preferably by reducing the need for developer-written code in the first place.

@@

This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to docuuments, replacing the hellscape of manual copy \& paste. But meeting this challenge may require a basic change of perspective in both live programming and local-first software.
Live programming must move from being a solitary activity to a collaborative workflow, and further where the collaboration is not just on editing source code but on live integrated code and data,
as in the Smalltalk/Lisp images of old.
%\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!}
Local-first software must move from automatically converging data replicas
to also interactively managing long-term divergence. That implies either application state
is being directly exposed to the user, or the application itself offers
version control affordances to the user. We encourage both communities to think outside the box of Git,
which has proven to baffle not only end-users but also a substantial fraction of developers. The Divergence challenge may be as much a technical challenge as a conceptual challenge, asking us to reconsider the entrenched boundaries between technologies in the stack.

@@

Although there is likely no silver bullet to navigate these tradeoffs, a good solution to this
problem would give developers or users tools to manage these tradeoffs. For example, a system
might allow developers to specify the desired behavior for a particular pair of schemas based
on the requirements of the domain. Or a system might even allow users to manually
disambiguate their intent for a given write.
Whatever the solution, it must be easy to learn and simple to use otherwise it won't be used in practice.

### 2024-08-24 revisions.tex
Clarify not a concrete tool
Reviewer A nitpicks

@@

We rewrote the introduction to remove statements of opinion and to clarify that the goal of our research on schema change is to speed feedback loops in software development. Ultimately that is a value judgement, but we believe that value is important to a substantial community of researchers and practitioners.

@@

@@

Reviewer A
-----------------------------------------------------------------
I will start from the end. In the conclusion you state: 'we developed a conceptual framework....' I disagree. You informally spoke about those ideas in interesting and tough provoking ways, but this is very far from developing a conceptual framework, and if I had to interpret your work as a conceptual framework I would find it severely lacking. However, I do not need to do so, I think your work can stand on its own as a set of considerations of those problems from a unifying viewpoint. This viewpoint is interesting, but is clearly limited and is basically only applicable to layered systems. With imperative systems I mean system with a set of predefined datatypes that are in some sense 'superior' to the user defined ones, that are just lists/records/unions of those primitive data types. This is not the case for example in lambda calculus, where there are no primitive data types, or in pure OO, where everything truly is an object and you could define string and ints as user defined types using programming patterns not dissimilar to church encoding.

Basically, as soon as 'true abstraction' is into place, most of your considerations stop being applicable.

"our work highlights the importance of.." No, you state that those things are important over and over, but you never make a good argument for it. In particular, you never ever state clearly -why it is important -what is it important for: that is, something is not good or bad in isolation. Something is good or bad instrumentally in order to reach a specific objective. What is the objective that you think is facilitated by having 'better more fluid programming environments'? You leave this implicit across the whole document!!

Section 1: "future programming systems will need to further..." Why? why will they need to do it? again, you do not "need to do" something in isolation. You need to do something in order to obtain or facilitate a goal. What is the goal here? I suspect once you state the goal I would be able to show you that it may not be the case.

So, "future programming systems will need to further..." is incomplete, the goal must be added, but even when the goal is added, this statement must be clearly labelled as an OPINION. I suspect it would be a widely agreed upon opinion, but an opinion; not a fact.

"will also need to transcend.." Again, add goal and flag as OPINION

"While we are firmly convinced that... are the future of programming" So much is implicit here. Do you think this is the future in the sense that you think is good and you think it will happen? Do you think it is the future because you hate it, but you still think it is going to be unavoidable? Do you think it will happen independently of your research work, or that it will happen because of your research work? There are 10 other possible interpretations for this sentence.

Just before 2.2 I think here you need to set expectations for correct local transformations, because no general solution exists, only conventions and expectations. Please, add numbers to those expectations, show examples of corner cases where stuff breaks and is an acceptable evil, and cases were stuff breaks on purpose (in some sense, and intended evil) To make an example: Java is of course not truly "cross-platform": if you wanted to put a formal definition to cross platform, the program should work identically in all environments. This would imply that the program should not be able to discover it's running environment. In Java I can make an 'if' at the beginning of the main and throw error on Windows. This makes Java not truly "cross-platform" in a logical sense, but this is often considered desirable.

In the context of those data transformations, there must be plenty of situations like that, where a change in behavior is not only possible or acceptable, but even desirable. However, we should have some clear formal definition of when/how/why such changes in behavior are introduced.

pg 7: "the key requirement.... do minimal harm" This is an Opinion + Industry convention.

"not commonly supported in data APIs" Define what a data API is and add a citation.

pg 10: "when applying code migrations, the system should minimize... the need for developer -written code" Again, "should" requires an explicit objective we are working toward. Plus, even after you add the objective, this still need to be labelled as an Opinion.

pg 13: Here I'm starting to wonder why you use this model at all. Once we accept that code can be changed and updated, the difference between data and code disappear. We could simply represent all data as hard coded instructions recreating such data. This may produce a much more cohesive mental model for the problem at hand. For example in this new model programs are able to 'restart' and when they restart they will execute this code 'pre-loading' the data in the state it was when the program was terminated.

Once code can be updated, data is just another kind of code. This requires a language where it is easy to declaratively describe any kind of arbitrary reachable object graphs, including circular ones.

Pg 16: "Live programming must move.... affordances" This is again an Opinion, and again 'must' can not exists alone, it needs a GOAL we are trying to obtain. You may think that such goal is really 'obvious', but it is not. What kind of programming you want to facilitate? why you want to facilitate that kind of programming? that kind of programming would in turn facilitate certain human endeavors but maybe not others, and the results may be very hard to predict and quite chaotic.

Pg 18: "no silver bullet to navigate... a good solution... tradeoffs" This assume no perverse incentives. Consider that if an operation is easier such operation will be done more often, and by sloppier and less skilled operators. Assuming the operators will be competent and will care about what they are doing is unrealistic.

>>>>

\section{Summary review}

We have clarified that this paper is not about a tool under development. Some of the problems are extracted from our different research projects but none of these projects span the rull range of problems.

\textrcolor{red}{explicitly speculative section?}

@@

We rewrote the introduction to clarify that the goal of our research on schema change is to speed feedback loops in software development. Ultimately that is a value judgement, but we believe that value is important to a substantial community of researchers and practitioners. We have qualified statements of opinion throughout the paper.

@@

We clarified that the goal of Divergence in End-User Workflows is to create a new kind of feedback channel for end users rather than speeding up existing channels.

In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of oferring solutions simple enough to be actually used.

@@
